---

title: "Reproducible Research: Peer Assessment 1"
author: "Jesús Pérez"
date: "7/30/2020"
output: 
  html_document:
    keep_md: true

---

# Activity monitoring

## 1. Loading and opening the data

```{r Open the file, echo = TRUE}

## Loading the file as a zip file
path <- getwd()
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, destfile = paste(path, "Files.zip", sep = "/"))
unzip(zipfile = "Files.zip")

## Open the data
data <- read.csv("activity.csv", header = TRUE)
str(data)
head(data)
```

Now we can see some parameters such as maximum, minimum, range and the NA's values of each variable using the **summary** function.

```{r summary1, echo = TRUE}
summary(data)
```

### 1.1 subsetting the data

Before plot the histogram we need to subset the data. For that purpose the **aggregate** function was used. The NA's values were desconsired.

```{r process data, echo = TRUE}

df <- aggregate(data$steps, by=list(date=data$date), FUN=sum)
df <- df[!is.na(df$x),]
summary(df)
head(df)
````

## 2. Plotting the Total per day histogram

To plot the histogram the **ggplot2** package was used. This is the ***histogram*** and not the ***barplot***. A histogram is used to continuous variables and shows the frequency of that particularly variable.
 
```{r histogram, echo = TRUE }
library(ggplot2)
p <- ggplot(df, aes(x=x))
p + geom_histogram(binwidth=1000, 
                   fill="#69b3a2", 
                   color="#e9ecef", 
                   alpha=0.7) + 
        labs(x= "Total steps per day", 
             y = "Frequency", 
             title = "Histogram of the total steps taken each day")

```

## 3. Mean and median 

The Mean and median number of steps taken each day can be obtained using the **summary** function (see section 7.1).

``` {r summary, echo = TRUE}
summary(df$x)
```
## 4. Time series plot of the average number of steps taken

For the second plot a new subsetting was made using the **aggregate** function, and then graphicated using **ggplot2**. 

``` {r plot2, echo = TRUE}
## Subsetting
df2 <- aggregate(data$steps, 
                 by=list(l=data$interval), 
                 FUN=mean, na.rm = TRUE)

## Calculating the max
maxx <- df2$l[which.max(df2$x)]

## Plotting 
p <-    ggplot(df2, aes(x = l, y = x)) +
        geom_line() + 
        labs(x= "Interval", y = "Average of steps taken", 
                title = "Average of steps taken by Time Interval") +
        geom_vline(xintercept= maxx, linetype="dashed", color = "red")
print(p)
```

## 5. Maximum number of steps by time interval

The 5-minute interval that, on average, contains the maximum number of steps was calculate using the **which.max** function (See plot 2) and the result was reported in hours.


``` {r maximum, echo = TRUE}
library(lubridate)

## Calculating the max
maxx <- df2$l[which.max(df2$x)]

## Transform to hours
duration(maxx, "minutes")
```

## 6. Imputing missing values

As there are days/intervals with missing values \textcolor{red}{NA} a new datasets was created. But, first the porcentage of \textcolor{red}{NA} was calculated. (Note that the total \textcolor{red}{NA} values were reported using the **summary** function in section **1**)

``` {r NA, echo = TRUE}
NAt <- sum(is.na(data$steps))
t <- nrow(data)
NAp <- (NAt/t)*100
NAp
```

### 6.1 Filling in the missing values

The \textcolor{red}{NA} values were filled with the mean values for their respective 5-minute interval using the **mutate** function of the **dyplr** package. A new dataset identically to the original but with the missing data filled in was created.

```` {r treating NA values, echo = TRUE}
library(dplyr)
df3 <- cbind(df2$x, data)
names(df3)[names(df3) == "df2$x"] <- "MeanSteps"
df4 <- df3 %>% mutate(MissingSteps = ifelse(is.na(steps), MeanSteps, steps))
df4 <- df4[, 2:5]
head(df4)

```

## 7 Histogram of the total number of steps taken each day after missing values

The new histogram was created with the new dataset. The results show that are no evident differences between plot 1 and 3, that means the ausence of \textcolor{red}{NA} values did not affect the analysis.

``` {r plot3, echo = TRUE}
df5 <- aggregate(df4$MissingSteps, by=list(date=df4$date), FUN=sum)
p <- ggplot(df5, aes(x=x))
p + geom_histogram(binwidth=1000, 
                   fill="#404080", 
                   color="#e9ecef", 
                   alpha=0.6) + 
        labs(x= "Total steps per day", 
             y = "Frequency", 
             title = "Histogram of the total steps taken each day without NA values")

````

### 7.1 New Mean and Median

To show the mean and median of the new dataset a table was created. The analysis shows that there were not differences between the original dataset and the dataset imputed with \textcolor{red}{NA} values. So, imputing missing data on the estimates of the total daily number of steps did not alterate the results.

``` {r mean and median comparisson, echo = TRUE}
Mean <- c(mean(df$x), mean(df5$x))
Median <- c(median(df$x), median(df5$x))
Comparisson <- cbind(Mean, Median)
Comparisson

```

## 8. Activity patterns between weekdays and weekends

To analyse the differences in the activity patterns between weekdays and weekends a new dataset was created. For this reason, the **weekdays** function was used, first converting the date to Date objects and then calculating the DayType as a factor. 

``` {r Day type, echo = TRUE}
df4$date <- as.Date(df4$date)
day <- as.factor(weekdays(df4$date))
df6 <- cbind(df4, day)
weekend <- c("Satruday", "Sunday")
df6 <- df6 %>% mutate(DayType = ifelse(day == weekend, "Weekend", "Weekday"))
df6$DayType <- as.factor(df6$DayType)
head(df6)
```

### 8.1 Panel plot

With the new dataset the results were plot using **ggplot2**. These plots clrearly show an important difference between the Activity patterns between weekdays and weekends.

``` {r plot4, echo = TRUE} 
df7 <- df6 %>% group_by(interval,DayType) %>% summarise(AverageSteps=mean(MissingSteps))
p <- df7 %>% ggplot(aes(x = interval, y = AverageSteps)) +
        geom_line() + 
        facet_grid(DayType ~ .) +
        labs(x= "Interval", y = "Average of steps taken by Time Interval", 
             title = "Average of steps taken by Time Interval and Day Type")
print(p)
```



